Beliefs about the development of mental life

### 2021-08-08 paper/paper.Rmd
tweaks to text

In this paper, we examine how adults conceptualize the developing minds of infants and children. We draw on techniques developed in studies of mind perception, in which researchers use large datasets and modern statistical methods to reveal conceptual representations of "the mind," broadly construed. In a groundbreaking study, Gray et al. asked participants to compare the mental capacities of a range of humans, animals, technologies, and other entities, and argued that participants’ judgments revealed a fundamental distinction between "agency" (e.g., capacities for self-control, moral reasoning, memory, and planning) and "experience" (e.g., capacities for hunger, fear, pain, and pleasure) [@gray.etal2007]. More recently, Weisman et al. used a similar empirical approach to argue that people’s conceptions of mental life are anchored by three fundamental components: "body" (physiological sensations such as hunger and pain), "heart" (social-emotional abilities such as embarrassment and pride), and "mind" (perceptual-cognitive capacities such as vision and memory [@weisman.etal2017a]); see [@malle2019] for a similar three-factor solution.

However, the burgeoning field of mind perception has largely neglected the question of how the mind emerges over development. In many studies, infants and children have simply not been included as targets for mental capacity attributions; instead, participants have been asked to reason about human adults alongside a range of non-human entities, including other animals, technologies, and supernatural beings. Even studies that have probed participants’ mental capacity attributions to infants and children have generally treated them as beings distinct from adults, rather than as snapshots of a single mind as it emerges and changes over time. For example, Gray et al.’s early work on mind perception suggested that a five-month-old infant is an entity that is generally considered to have experience but not agency—in contrast to, say, a five-year-old child (considered to have experience as well as limited agency), a human adult (considered to have maximal experience and agency), or a robot (considered to have agency but not experience) [@gray.etal2007]. Likewise, Weisman et al.’s studies characterized infants and young children—along with, e.g., chimpanzees, elephants, and dolphins—as entities that are perceived to be fully capable of physiological sensations but more limited in their perceptual-cognitive capacities and social-emotional abilities; this is in contrast to human adults, who are perceived to be fully capable across these domains [@weisman.etal2017a].

Although such findings can be reinterpreted as providing preliminary evidence that people consider the minds of infants and children to differ from the minds of adults, they do not address how people conceptualize the development of the mind in early life.

Of course, infants and young children are far from unusual social partners. Caregiving relationships are at the core of human existence; for many people, infants and young children are among the most frequent and highly valued minds that we encounter. Caring for a young child involves making countless decisions under uncertainty: Why is my child crying, smiling, shouting, or biting? Can they understand what I am saying? Does my child love me? As in other domains of reasoning [@gerstenberg.tenenbaum2017; @gopnik.wellman1994; @wellman.gelman1992], people likely draw on folk beliefs and intuitive theories about how the mind develops to predict, explain, and respond to children’s behavior and development, with consequences for the well-being of children in their care.

@@

What does it mean that these four factors emerged? At a most basic level, our results reflect sets of intercorrelated capacities: When a participant judged that a child of a certain age was highly capable of planning, they also tended to judge that that child was capable of having self control, thinking before they act, and the other capacities with strong factor loadings on the _cognition and control_ factor; and likewise for the suites of capacities that define the other factors revealed by EFA. Therefore, we take these factors to reflect latent constructs—fundamental components of mental life relevant to reasoning about human children [@weisman.etal2017a]. Moreover, because each participant was asked to reason not just about some class of target characters (e.g., "chimpanzees," "robots"), but about the human child over development, we further propose that part of what contributed to the differentiation of these four factors was participants’ perceptions of development: the extent to which different capacities are present at birth, the rate at which different capacities develop, and the mechanisms that drive this development. In Studies 2 and 3, we explore these possibilities in greater depth.

@@

In terms of the covariance structure of participants' capacity ratings, a preregistered EFA yielded four factors very similar to the four factors from Study 1 (see Supplemental Materials), with all capacity items loading strongly on the factors that they were selected to represent. Minimally, these results suggest that our selection of mental capacities for this study captured the essential meaning of the four factors revealed in Study 1. Beyond this, we consider these results to be a somewhat independent conceptual replication of Study 1 results and validation of our interpretation of these factors. After all, given that Study 2 featured only 20 of the 60 capacities featured in Study 1, it certainly could have been the case that this analysis would suggest retaining fewer factors (as often occurs with fewer variables). Conversely, given that Study 2 featured 13 within-subjects observations for each capacity compared to the 3 in Study 1, this analysis could have surfaced more than four factors (as often occurs with more observations). Finally, if we had misinterpreted the fundamental semantic content that distinguished the factors in Study 1 and selected inappropriate capacities to represent these factors, the Study 2 analysis could have identified more capacities with cross-loadings. Instead, we believe the results of this analysis offer some validation of the four factors we have called _bodily sensations_, _negative affect_, _social connection_, and _cognition and control_.

@@

Regressing participants' item-level responses onto target age, domain, and interactions between them (via a multilevel generalized additive model; see Methods) confirmed both of our preregistered hypotheses: On average, participants rated older children as more capable, but perceptions of the development of children’s capacities differed dramatically across domains. On one extreme, participants attributed some, but relatively little, capacity for _cognition and control_ to newborns (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(sd) %>% round_x(., 2)`), but reported that such capacities increase dramatically over the target age range (average smoothing term, analogous to a slope in linear regression: $F$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorCognition and control", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`). Compared to these attributions of _cognition and control_, participants attributed substantially greater capacities to newborns in the domains of _social connection_ (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s2_gam_ptab["factorsoc_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorsoc_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorsoc_cog", "Pr(>|t|)"], 3)))`) and _negative affect_ (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds: $\beta$ = `r round_x(s2_gam_ptab["factorneg_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorneg_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorneg_cog", "Pr(>|t|)"], 3)))`) and, accordingly, reported relatively small increases over the target age range in both domains (average smoothing terms for _social connection_: $F$ = `r round_x(s2_gam_stab["s(target_year):factorSocial connection", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorSocial connection", "p-value"] < 0.001, "< 0.001", paste("=",  round_x(s2_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorSocial connection", "edf"], 2)`; for _negative affect_: $F$ = `r round_x(s2_gam_stab["s(target_year):factorNegative affect", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorNegative affect", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorNegative affect", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`). At the other extreme, in the domain of _bodily sensations_ participants attributed nearly maximal capacity to newborns (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s2_gam_ptab["factorbod_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorbod_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorbod_cog", "Pr(>|t|)"], 3)))`) and a significant but small increase in capacities for _bodily sensations_ over the target age range (average smoothing term: $F$ = `r round_x(s2_gam_stab["s(target_year):factorBodily sensations", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorBodily sensations", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorBodily sensations", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorBodily sensations", "edf"], 2)`). See Figure \@ref(fig:fig02), Panel A.

Visual inspection of Figure \@ref(fig:fig02), Panel A clearly illustrates that the shape of perceived growth in children’s capacities varied substantially across domains. In the domain of _bodily sensations_, participants perceived hardly any growth over the target age range, and what growth they did perceive occurred in the first year of life. Participants perceived development in the domains of _negative affect_ and _social connection_ to be quite non-linear, characterized by rapid development over the first 1-2 years of life, at which point the curves level off dramatically. In contrast, participants perceived dramatic development in the domain of _cognition and control_, but perceived this development to unfold fairly steadily across the full age range.

@@

The results of Study 3 yielded a very similar picture of US adults’ understanding of the development of human mental life; see Figure \@ref(fig:fig02), Panel B. As in Study 2, participants perceived that newborns have very little capacity for _cognition and control_ (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(sd) %>% round_x(., 2)`), but reported that such capacities increase dramatically over the target age range (average smoothing term: $F$ = `r round_x(s3_gam_stab["s(target_year):factorCognition and control", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorCognition and control", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`). Again, participants attributed substantially greater capacities to newborns in the domains of _social connection_ (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds: $\beta$ = `r round_x(s3_gam_ptab["factorsoc_cog", "Estimate"], 2)`, $p$ `r ifelse(s3_gam_ptab["factorsoc_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_ptab["factorsoc_cog", "Pr(>|t|)"], 3)))`) and _negative affect_ (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds: $\beta$ = `r round_x(s3_gam_ptab["factorneg_cog", "Estimate"], 2)`, $p$ `r ifelse(s3_gam_ptab["factorneg_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_ptab["factorneg_cog", "Pr(>|t|)"], 3)))`); in Study 3, as in Study 2, participants reported relatively small increases over the target age range in the domain of _negative affect_ (average smoothing term: $F$ = `r round_x(s3_gam_stab["s(target_year):factorNegative affect", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorNegative affect", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_stab["s(target_year):factorNegative affect", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`), but in this case they reported quite dramatic increases in the domain of _social connection_ ($F$ = `r round_x(s3_gam_stab["s(target_year):factorSocial connection", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorSocial connection", "p-value"] < 0.001, "< 0.001", paste("=",  round_x(s3_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorSocial connection", "edf"], 2)`). Finally, as in Study 2, in the domain of _bodily sensations_ participants attributed nearly maximal capacity to newborns (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s3_gam_ptab["factorbod_cog", "Estimate"], 2)`, $p$ `r ifelse(s3_gam_ptab["factorbod_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_ptab["factorbod_cog", "Pr(>|t|)"], 3)))`) and a significant but small increase in capacities for _bodily sensations_ over the target age range (average smoothing term: $F$ = `r round_x(s3_gam_stab["s(target_year):factorBodily sensations", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorBodily sensations", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_stab["s(target_year):factorBodily sensations", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorBodily sensations", "edf"], 2)`). See Figure \@ref(fig:fig02), Panel B.

Visual inspection of Figure \@ref(fig:fig02), Panel B makes it clear that the shape of these perceived developmental trajectories varied substantially across domains, echoing quite precisely the results of our post-hoc exploratory analyses in Study 2. In the domain of _bodily sensations_, participants perceived hardly any growth over the target age range, and what growth they did perceive was understood to occur rapidly in the first year of life. In the domains of _negative affect_ and _social connection_, participants perceived much more growth over the target age range, and in both cases most of this growth was understood to occur during the first two years of life; again, this was especially pronounced in the domain of _social connection_. Finally, participants perceived development in the domain of _cognition and control_ to be much more gradual and, by extension, more linear across the target age range.

See Supplemental Materials for the full results of this analysis.

@@

Why do people believe certain aspects of mental life, but not others, to be present at birth? What do people perceive to be driving the development of capacities for _bodily sensations_, _negative affect_, _social connection_, and _cognition and control_? As a first step in what we hope to be a longer line of work addressing such questions, we asked participants to assess the importance of a variety of possible developmental mechanisms for each of the capacities included in this study. We theorized that when people perceive that a capacity is substantially present at birth, they believe that the capacity is innate, and, conversely, that when people perceive that a capacity develops slowly over years, they believe the capacity is in some sense learned. (See [@wang.feigenson2019; @berent.etal2019; @berent.etal2020] for in-depth investigations of ordinary people’s intuitions about the innateness of cognitive abilities, motor abilities, and emotion.)

@@

In a series of three large-scale studies, we identified four distinct suites of mental capacities that are perceived by US adults to develop over human infancy and early childhood, _bodily sensations_, _negative affect_, _social connection_, and _cognition and control_ (Study 1); charted how these different aspects of mental life are perceived to change over the first five years of a child’s life (Studies 2-3); and explored the lay theories that underlie the observed differences in the perceived developmental trajectories of these four aspects of mental life (Study 3).

>>>>

In this paper, we examine how adults conceptualize the developing minds of infants and children. We draw on techniques developed in studies of mind perception, in which researchers use large datasets and modern statistical methods to reveal conceptual representations of "the mind," broadly construed. In a groundbreaking study, Gray et al. asked participants to compare the mental capacities of a range of humans, animals, technologies, and other entities, and argued that participants’ judgments revealed a fundamental distinction between "agency" (e.g., capacities for self-control, moral reasoning, memory, and planning) and "experience" (e.g., capacities for hunger, fear, pain, and pleasure) [@gray.etal2007]. More recently, Weisman et al. used a similar empirical approach to argue that people’s conceptions of mental life are anchored by three fundamental components: "body" (physiological sensations such as hunger and pain), "heart" (social-emotional abilities such as embarrassment and pride), and "mind" (perceptual-cognitive capacities such as vision and memory [@weisman.etal2017a]; see [@malle2019] for a similar three-factor solution. Such findings have inspired a wide range of subsequent work on topics as wide-ranging as moral reasoning, dehumanization and stereotyping, beliefs about the afterlife, and human-robot interactions.

However, the burgeoning field of mind perception has largely neglected the question of how the human mind emerges over development. In many studies, infants and children have simply not been included as targets for mental capacity attributions; instead, participants have been asked to reason about human adults alongside a range of non-human entities, including other animals, technologies, and supernatural beings. Even studies that have probed participants’ mental capacity attributions to infants and children have generally treated them as beings distinct from adults, rather than as snapshots of a single mind as it emerges and changes over time. For example, Gray et al.’s early work on mind perception suggested that a five-month-old infant is an entity that is generally considered to have experience but not agency—in contrast to, say, a five-year-old child (considered to have experience as well as limited agency), a human adult (considered to have maximal experience and agency), or a robot (considered to have agency but not experience) [@gray.etal2007]. Likewise, Weisman et al.’s studies characterized infants and young children—along with, e.g., chimpanzees, elephants, and dolphins—as entities that are perceived to be fully capable of physiological sensations but more limited in their perceptual-cognitive capacities and social-emotional abilities; this is in contrast to human adults, who are perceived to be fully capable across these domains [@weisman.etal2017a]. Although such findings can be reinterpreted as providing preliminary evidence that people consider the minds of infants and children to differ from the minds of adults, they do not address how people conceptualize the development of the mind in early life.

Yet infants and young children are far from unusual social partners. Caregiving relationships are at the core of human existence; for many people, infants and young children are among the most frequent and highly valued minds that we encounter. Caring for a young child involves making countless decisions under uncertainty: Why is my child crying, smiling, shouting, or biting? Can they understand what I am saying? Does my child love me? As in other domains of reasoning [@gerstenberg.tenenbaum2017; @gopnik.wellman1994; @wellman.gelman1992], people likely draw on folk beliefs and intuitive theories about how the mind develops to predict, explain, and respond to children’s behavior and development, with consequences for the well-being of children in their care.

@@

What does it mean that these four factors emerged? At a most basic level, our results reflect sets of intercorrelated capacities: When a participant judged that a child of a certain age was highly capable of planning, they also tended to judge that that child was capable of having self control, thinking before they act, and the other capacities with strong factor loadings on the _cognition and control_ factor; and likewise for the suites of capacities that define the other factors revealed by EFA. Therefore, we take these factors to reflect latent constructs—fundamental components of mental life relevant to reasoning about human children. Moreover, because each participant was asked to reason not just about some class of target characters (e.g., "chimpanzees," "robots"), but about the human child over development, we further propose that part of what contributed to the differentiation of these four factors was participants’ perceptions of development: the extent to which different capacities are present at birth, the rate at which different capacities develop, and the mechanisms that drive this development. In Studies 2 and 3, we explore these possibilities in greater depth.

@@

A preregistered EFA yielded four factors very similar to the four factors from Study 1, with all capacity items loading strongly on the factors that they were selected to represent and less strongly on other factors (see Supplemental Materials). Minimally, this suggests that our selection of mental capacities for this study captured the essential meaning of the four factors revealed in Study 1. Beyond this, we consider these results to be a somewhat independent conceptual replication of Study 1 results and validation of our interpretation of these factors. After all, given that Study 2 featured only 20 of the 60 capacities featured in Study 1, it certainly could have been the case that this analysis would suggest retaining fewer factors (as often occurs with fewer variables). Conversely, given that Study 2 featured 13 within-subjects observations for each capacity compared to the 3 in Study 1, this analysis could have surfaced more than four factors (as often occurs with more observations). Finally, if we had misinterpreted the fundamental semantic content that distinguished the factors in Study 1 and selected inappropriate capacities to represent these factors, the Study 2 analysis could have identified more capacities with cross-loadings. Instead, the results of this analysis offer some validation of the four factors we have called _bodily sensations_, _negative affect_, _social connection_, and _cognition and control_.

@@

Regressing participants' item-level responses onto target age, domain, and interactions between them via a multilevel generalized additive model confirmed both of our preregistered hypotheses: On average, participants rated older children as more capable, but perceptions of the development of children’s capacities differed dramatically across domains. On one extreme, participants attributed some, but relatively little, capacity for _cognition and control_ to newborns (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(sd) %>% round_x(., 2)`), but reported that such capacities increase dramatically over the target age range (average smoothing term, analogous to a slope in linear regression: $F$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorCognition and control", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`). Compared to these attributions of _cognition and control_, participants attributed substantially greater capacities to newborns in the domains of _social connection_ (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s2_gam_ptab["factorsoc_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorsoc_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorsoc_cog", "Pr(>|t|)"], 3)))`) and _negative affect_ (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds: $\beta$ = `r round_x(s2_gam_ptab["factorneg_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorneg_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorneg_cog", "Pr(>|t|)"], 3)))`) and, accordingly, reported relatively small increases over the target age range in both domains (average smoothing terms for _social connection_: $F$ = `r round_x(s2_gam_stab["s(target_year):factorSocial connection", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorSocial connection", "p-value"] < 0.001, "< 0.001", paste("=",  round_x(s2_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorSocial connection", "edf"], 2)`; for _negative affect_: $F$ = `r round_x(s2_gam_stab["s(target_year):factorNegative affect", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorNegative affect", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorNegative affect", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`). At the other extreme, in the domain of _bodily sensations_ participants attributed nearly maximal capacity to newborns (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s2_gam_ptab["factorbod_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorbod_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorbod_cog", "Pr(>|t|)"], 3)))`) and a significant but small increase in capacities for _bodily sensations_ over the target age range (average smoothing term: $F$ = `r round_x(s2_gam_stab["s(target_year):factorBodily sensations", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorBodily sensations", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorBodily sensations", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorBodily sensations", "edf"], 2)`).

Visual inspection of Figure \@ref(fig:fig02), Panel A, clearly illustrates that the shape of perceived growth in children’s capacities varied substantially across domains. In the domain of _bodily sensations_, participants perceived hardly any growth over the target age range, and what growth they did perceive occurred in the first year of life. Participants perceived development in the domains of _negative affect_ and _social connection_ to be quite non-linear, characterized by rapid development over the first 1-2 years of life, at which point the curves level off dramatically. In contrast, participants perceived dramatic development in the domain of _cognition and control_, but perceived this development to unfold fairly steadily across the full age range.

@@

This approach yielded a very similar picture of US adults’ understanding of the development of human mental life; see Figure \@ref(fig:fig02), Panel B. As in Study 2, participants perceived that newborns have very little capacity for _cognition and control_ (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(sd) %>% round_x(., 2)`), but reported that such capacities increase dramatically over the target age range (average smoothing term: $F$ = `r round_x(s3_gam_stab["s(target_year):factorCognition and control", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorCognition and control", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`). Again, participants attributed substantially greater capacities to newborns in the domains of _social connection_ (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds: $\beta$ = `r round_x(s3_gam_ptab["factorsoc_cog", "Estimate"], 2)`, $p$ `r ifelse(s3_gam_ptab["factorsoc_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_ptab["factorsoc_cog", "Pr(>|t|)"], 3)))`) and _negative affect_ (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds: $\beta$ = `r round_x(s3_gam_ptab["factorneg_cog", "Estimate"], 2)`, $p$ `r ifelse(s3_gam_ptab["factorneg_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_ptab["factorneg_cog", "Pr(>|t|)"], 3)))`); in Study 3, as in Study 2, participants reported relatively small increases over the target age range in the domain of _negative affect_ (average smoothing term: $F$ = `r round_x(s3_gam_stab["s(target_year):factorNegative affect", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorNegative affect", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_stab["s(target_year):factorNegative affect", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`), but in this case they reported quite dramatic increases in the domain of _social connection_ ($F$ = `r round_x(s3_gam_stab["s(target_year):factorSocial connection", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorSocial connection", "p-value"] < 0.001, "< 0.001", paste("=",  round_x(s3_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorSocial connection", "edf"], 2)`). Finally, as in Study 2, in the domain of _bodily sensations_ participants attributed nearly maximal capacity to newborns (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s3_gam_ptab["factorbod_cog", "Estimate"], 2)`, $p$ `r ifelse(s3_gam_ptab["factorbod_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_ptab["factorbod_cog", "Pr(>|t|)"], 3)))`) and a significant but small increase in capacities for _bodily sensations_ over the target age range (average smoothing term: $F$ = `r round_x(s3_gam_stab["s(target_year):factorBodily sensations", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorBodily sensations", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_stab["s(target_year):factorBodily sensations", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorBodily sensations", "edf"], 2)`).

Visual inspection of Figure \@ref(fig:fig02), Panel B, makes it clear that the shape of these perceived developmental trajectories varied substantially across domains, echoing quite precisely the results of our post-hoc exploratory analyses in Study 2.

@@

Why do people believe certain aspects of mental life, but not others, to be present at birth? What do people perceive to be driving the development of capacities for _bodily sensations_, _negative affect_, _social connection_, and _cognition and control_? As a first step in what we hope to be a longer line of work addressing such questions, we asked participants to assess the importance of a variety of possible developmental mechanisms for each of the capacities included in this study. We theorized that when people perceive that a capacity is substantially present at birth, they believe that the capacity is innate, and, conversely, that when people perceive that a capacity develops slowly over years, they believe the capacity is in some sense learned. (See @wang.feigenson2019; @berent.etal2019; @berent.etal2020 for in-depth investigations of ordinary people’s intuitions about the innateness of cognitive abilities, motor abilities, and emotion.)

@@

In a series of three large-scale studies, we identified four distinct suites of mental capacities that are perceived by US adults to develop over human infancy and early childhood, _bodily sensations_, _negative affect_, _social connection_, and _cognition and control_ (Study 1); measured perceptions of these capacities at birth and charted how they are perceived to change over the first five years of a child’s life (Studies 2-3); and explored the lay theories that underlie the observed differences in the perceived developmental trajectories of these four aspects of mental life (Study 3).
